Automatic discovery of circuit structure from neural connectomics

### 2014-01-06 docs/paper.tex
konrad updates

\begin{abstract}
  New techniques produce massive data about neural connectivity,
  necessitating new analysis methods to discover the biological and
  computational basis of this connectivity. It has long been assumed
  that discovering the local patterns of microcircuitry is the key to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. We show that the approach recovers
  known neuron types in the retina, reveals interesting structure in
  the nervous system of c. elegans, and automatically discovers the
  structure of microprocessors. Our approach extracts structural
  meaning from connectomics, enabling new approaches of deriving
  anatomical insights from these emerging datasets.

@@

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity patterns
(microcircuits) has been prominent in many areas, ranging from sensory
(e.g. retina, \autocite{Masland2001} via processing (e.g. neocortex
\autocite{Mountcastle1997}) to movement (e.g. spinal cord)
\autocite{Grillner2005}. Indeed, such repeated patterns also exists in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to back
out underlying microcircuitry.

@@

\section{Results}
We use a Bayesian structure discovery approach that is based on prior
knowledge about the way positions and cell types affect the
probability of connections between cells, a so-called generative
model. Basing our model on previous infinite stochastic block models
(iSBM)\autocite{Kemp2006a,Xu2006}, we assume that each neuron is a
member of a single ``type'', and that the probability of two neurons
being connected is a function of their types. Crucially, we
incorporate the prior knowledge that spatial proximity matters, adding
distance-dependence to the iSBM.  For example, retinal ganglion cells
may synapse on nearby, but not far away, amacrine cells.  Note that
this simplistic prior information is all that we tell our model; no
additional physiological information is included.  We term our model
the ``infinite spatial-relational model'' (iSRM), in keeping with the
existing nonparametric literature, as it discovers a potentially
unbounded number of unobserved, or ``latent'', cell types.

We combine annealing with Markov Chain Monte Carlo to
perform simultaneously joint posterior inference over the structure,
parameters, and hyperparameters of the model (see Methods). This
procedure automatically learns the number of cell types, which cells 
belong to which types, along with the 
the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). These results can then be compared with the
results of previous anatomical studies that have used other aspects of
neurons, such as chemical properties and cell-morphology to divide
neurons into types.

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons deterministically known. This
incredible cross-organism consistency led to early mapping of the
c. elegans chemical and electrical (gap junction) connectome. Unlike
the retina, only the motor neurons in c. elegans exhibit regular
distribution in space, being distributed along the body axis. Most
interneurons are concentrated in various ganglia that project
throughout the entire animal, and the sensory neurons are primarily
located in a small number of anterior ganglia.

>>>>

\begin{abstract}
  New techniques produce massive data about neural connectivity,
  necessitating new analysis methods to discover the biological and
  computational basis of this connectivity. It has long been assumed
  that discovering the local patterns of microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. We show that the approach recovers
  known neuron types in the retina, reveals interesting structure in
  the nervous system of c. elegans, and automatically discovers the
  structure of microprocessors. Our approach extracts structural
  meaning from connectomics, enabling new approaches of deriving
  anatomical insights from these emerging datasets.

@@

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity patterns
(microcircuits) has been prominent in many areas, ranging from sensory
(e.g. retina, \autocite{Masland2001} via processing (e.g. neocortex
\autocite{Mountcastle1997}) to movement (e.g. spinal cord)
\autocite{Grillner2005}. Indeed, such repeated patterns also exists in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to automatically back
out underlying microcircuitry.

@@

\section{Results}
From the generic assumption that cell types exist and that connection probability depends on the involved cell types and distance, we develop a Bayesian structure discovery algorithm.
We start with existing infinite stochastic block models (iSBM)\autocite{Kemp2006a,Xu2006}, which have been shown to meaningfully cluster connection graphs. We extend this approach by adding distance dependence to model salient aspects of microcircuitry. For example, retinal ganglion cells may synapse on nearby, but not far away, amacrine cells.  We term our model the ``infinite spatial-relational model'' (iSRM), in keeping with the existing nonparametric literature. This procedure automatically learns the number of cell types, which cells belong to which types, along with the the distance and type dependent connectivity probabilities (see fig~\ref{fig:overview}). We apply our algorithm to datasets from mouse retina, c-elegans and a historical microprocessor. Anatomists classify cells based on many features, giving us a meaningful baseline to compare against.

@@

\subsection{C. elegans}
[TODO Say this is harder than retina because of these facts]
The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons deterministically known. This
incredible cross-organism consistency led to early mapping of the
c. elegans chemical and electrical (gap junction) connectome. Unlike
the retina, only the motor neurons in c. elegans exhibit regular
distribution in space, being distributed along the body axis. Most
interneurons are concentrated in various ganglia that project
throughout the entire animal, and the sensory neurons are primarily
located in a small number of anterior ganglia.
